Implementation of a minimal branch-decomposition algorithm for planar graphs.

### 2024-06-10 paper/main.tex
Update paper

\title{Implementation of a minimal branch-decomposition algorithm for simple planar graphs.}
\author{Kristoffer Højelse}
\date{February 2024}

@@

@@

\section{Motivation}
	Some graph optimization problems can be solved efficiently for graphs of small branchwidth.\ref{}

@@

@@

	A graph is called \textbf{simple} if and only if it has no parallel edges and no self-loops.

	A graph is called \textbf{undirected} if and only if all its edges can be traversed in both directions.

	A graph is called \textbf{connected} (or 1-vertex-connected) if and only if there exists a path between any two vertices.

	A graph is called \textbf{planar} if and only if there is a way to draw the graph in 2 dimensions such that no pair of edges crosses.

	A \textbf{branch decomposition} $B$ of a graph $G$ is a tree where every edge of $G$ is a leaf in $B$ and every internal vertex of $B$ has exactly 3 neighbors. $B$ is an unrooted binary tree.

	A \textbf{minimal branch decomposition} of $G$ is any branch decomposition of $G$ of minimal width among all branch decompositions of $G$.

@@

	The algorithm described in this paper solves \textsc{The Planar Minimal Branch Decomposition Problem} which can be computed in polynomial time.

@@

\section{The algorithm}
	This section describes the algorithm given by Seymour and Thomas\ref{} by identifying a set of practical problems and subproblems and how they relate.

@@

	Informally; The medial graph is a graph where there is a node in the medial for each edge, and an edge between two nodes if their corresponding edges are consecutive in some face of the graph.

	\begin{definition}
		The medial graph $M(G)$ of a connected plane graph $G$ is a graph with a vertex $e^*$ for each edge $e$ of $G$ and for each face $f$ of $G$, there's an edge $d^*$ between a pair of vertices $e_1^*$, $e_2^*$ of $M(G)$ if $e_1$ and $e_2$ are consecutive in $f$.
	\end{definition}

@@

	Doing a series of edge contractions (contraction of all edges between a pair of vertices) on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices can be assembled into a minimal carving decomposition of $M$.

@@

	\begin{problem}\label{problem:contraction}
		Given a graph $M$ that might have parallel edges, output a graph resulting from a contraction of all edges between a pair of vertices.
	\end{problem}

@@

	\begin{definition}
		Contraction.

		Given an undirected Graph $G=\{V,E\}$ with no self-loops and pair of vertices $u,v \in V$ such that $\{u,v\} \in E$, remove all edges between $u$ and $v$ and update every edge $\{v,w\} \in E$ to be $\{u,w\}$.
	\end{definition}

@@

	$E(Q(M, k, e)) = \{e_1 \mid$ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

@@

	\begin{definition}
		The dual graph $M^*$ of a plane graph $M$ is a graph where a face $f$ of $M$ is a vertex $f^*$ of $M^*$ and an edge $e$ that separates two faces $f_1$,$f_2$ of $M$ is an edge $e^*$ of $M^*$ between $f_1^*$ and $f_2^*$.
	\end{definition}

	\begin{corollary}
		If multiple edges separate $f_1$ and $f_2$ there will be parallel edges between $f_1*$ and $f_2*$.
	\end{corollary}

	\begin{corollary}\label{def:dual_coro_loop}
		If the same face is on both sides of $e$, then $e^*$ will be a self-loop.
	\end{corollary}

	For the algorithm in this paper, the class of graphs that will be given as input is medial graphs of simple undirected connected planar graphs. Therefore I claim \ref{claim:irri}.

	\begin{claim}\label{claim:irri}
		Corollary \ref{def:dual_coro_loop} will be irrelevant for any implementation of the algorithm.
	\end{claim}

@@

	Let $S$ be every possible state when the rat-catcher is in a face some of which might be losing states. $S = \{ (f, v) \mid v \in V(M) \land f$ is a face of $ M \}$.

	Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \mid e \in E(M) \land C$ is a connected component of $Q(M, k, e) \}$.

@@

		As this contraction is a contraction of ALL edges between a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the ordering invariant could be difficult, but luckily in this context, it is irrelevant.

@@

		First, update any edges incident to either $a$ or $b$. Then creating the neighborhood of the new vertex $c$ from the contraction of vertices $a$ and $b$, is done by firstly finding any shared edge $e$. In this implementation the first shared edge $e$ in the neighborhood of $a$. This edge has some ID $e$ and the other half-edge with ID $-e$ will therefore be in the neighborhood of $b$. Now  "rotating" the neighborhoods of $a$ and $b$ such edge $e$ and $-e$ is at index 0 in both lists means that a concatenation of the lists will preserve the ordering around the new vertex $c$. And finally, remove any edges between $a$ and $b$.

@@

		Considering a tuple $(f, v)$, if there exists a tuple $(e, C)$ which is losing where $e \in f$ and $v \in V(C)$ then $(f, v)$ is losing.

@@

		\begin{claim}
			The shortest closed walk that includes both $e_1^*$ and $e_2^*$ has the same length as either $$d(s_1, s_2) + d(t_1, t_2) + 2$$ or $$d(s_1, t_1) + d(s_2, t_2) + 2$$. Where $d(u, v)$ is the length of the shortest path from $u$ to $v$.
		\end{claim}

@@

		Using the mapping from links of the dual to edges, and the fact that an edge $e$ is called quiet iff. $e$ is not noisy, the quiet edges can be obtained in the natural way.

		Computing the quiet subgraph and the connected components thereof is done with a depth-first search approach.

		The edges of the connected components are irrelevant for the rest of the algorithm, so only a list of vertices is returned for each connected component.

@@

		The dual has a vertex for each face of the input graph. The faces are found by selecting an unmarked half-edge, and then marking all the edges of the face it belongs to, by following the edges which are just next to each other in the ordered neighborhoods.

@@

\section{Appendix.}
	\pyfile{branch_decomposition.py}{0}{1000}
	\pyfile{branch_width_brute_force.py}{0}{1000}	
	\pyfile{branch_width.py}{0}{1000}
	\pyfile{carving_decomposition.py}{0}{1000}
	\pyfile{carving_width_brute_force.py}{0}{1000}
	\pyfile{carving_width.py}{0}{1000}
	\pyfile{contraction.py}{0}{1000}
	\pyfile{dual_graph.py}{0}{1000}
	\pyfile{Graph.py}{0}{1000}
	\pyfile{medial_graph.py}{0}{1000}
	\pyfile{parse_graph.py}{0}{1000}
	\pyfile{parse_newick.py}{0}{1000}

>>>>

\title{Implementation of a minimal branch-decomposition algorithm for planar graphs.}
\author{Kristoffer Højelse}
\date{February 2024}

@@

% west https://dwest.web.illinois.edu/grammar.html
% The abstract states the results as fully as possible in a brief presentation. Crucial specialized terms the reader needs to know to understand the statements should be defined. The abstract stands on its own, especially in the age of electronic communication where it may be separate from the rest of the paper, and hence it contains no numbered reference to the bibliography.

@@

% The first section of the paper is an "Introduction" that should motivate the problem, discuss related results, state the results more completely, and perhaps summarize the techniques or the structure of the paper or crucial definitions.

\section{Introduction}
	Some graph optimization problems can be solved efficiently for graphs of small branchwidth.\cite{CNP+11}

	Pino\cite{Pin16} applies branch decompositions.

	Seymour and Thomas\cite{ST93} give the rat-catching algorithm.

	Bian, Gu and Zhu\cite{BGZ15} describe and benchmark some implementations.

@@

\section{Preliminaries}

	A \textit{graph} $G$ is a 3-tuple $G = (V, E, \psi)$ consisting of a vertex set $V$, an edge set $E$ and a function $\psi$, where $V \subset \mathbb{N}^+$ and where $E \subset \mathbb{N}^+$ and where $\psi \st E \to \{\{u,v\} \st u,v \in V \land u \neq v \}$.

	\textit{Note that other authors might instead name this an Undirected Labelled Multigraph (not permitting self-loops) (edges having own identity).}

	A graph $G$ is \textit{simple}, if it has no parallel edges; $\neg \exists e_1,e_2 \in E_G$ where $e_1 \neq e_2$ such that $\psi_G(e_1) = \psi_G(e_1)$.

	A \textit{subgraph} $H$ of a graph $G$, is a graph where some vertices and edges might be missing; a graph where $V_H \subseteq V_G$ and where $E_H \subseteq E_G$ and where $\psi_H$ is a subfunction of $\psi_G$; if $e \in E_H$ then $\psi_H(e) = \psi_G(e)$.

	A graph is \textit{connected} if there exists a path connecting any two vertices.

	A \textit{component} of a graph, is a connected subgraph.

	A graph is \textit{planar} if there exists a planar embedding of the graph.

	A \textit{plane graph} (syn. \textit{planar embedding of the graph}), is a 2D drawing of a planar graph such that no edges are crossing. In other words, for a planar graph $G$ a \textit{plane graph}, is a set of points $P$ and line segments $L$, with a one-to-one correspondence between $V(G)$ and $P$ and with a one-to-one correspondence between $E(G)$ and $L$, such that one of the endpoints of the line segments for the corresponding edges that are incident to some vertex $v$ occupy the same coordinate as the point corresponding to $v$ and such that line segments only pairwise intersect at their endpoints.

	\begin{definition}
		(\textit{Contraction})

		Given an graph $G$ and pair of vertices $u,v \in G_V$ such that $\{u,v\}$ is in the image of $\psi_G$, remove all edges incident to $u$ and $v$ and update every edge $\{v,w\} \in E$ to be $\{u,w\}$.
	\end{definition}

	\begin{definition}
		(\textit{Dual Graph})

		The dual graph $M^*$ of a plane graph $M$ is a graph where a face $f$ of $M$ is a vertex $f^*$ of $M^*$ and an edge $e$ that separates two faces $f_1$,$f_2$ of $M$ is an edge $e^*$ of $M^*$ incident to $f_1^*$ and $f_2^*$.
	\end{definition}

	\begin{corollary}
		If multiple edges separate $f_1$ and $f_2$ there will be parallel edges incident to $f_1*$ and $f_2*$.
	\end{corollary}

	\begin{definition}
		(\textit{Medial Graph})

		The medial graph $M(G)$ of a connected plane graph $G$ is a graph with a vertex $e^*$ for each edge $e$ of $G$ and for each face $f$ of $G$, there's an edge $d^*$ incident to a pair of vertices $e_1^*$, $e_2^*$ of $M(G)$ if $e_1$ and $e_2$ are consecutive in $f$.
	\end{definition}

	\begin{corollary}
		A medial graph is a 4-regular plane graph.
	\end{corollary}

	A \textit{Branch Decomposition} $B$ of a graph $G$ is a tree where every edge of $G$ is a leaf in $B$ and every internal vertex of $B$ has exactly 3 neighbors. $B$ is an unrooted binary tree.

	A \textit{Minimal Branch Decomposition} of $G$ is any branch decomposition of $G$ of minimal width among all branch decompositions of $G$.

	Let $n = |V(G)|$ and $m = |E(G)|$.

	A \textit{Carving Decomposition}

	A \textit{Minimal Carving Decomposition}

@@

@@

	The algorithm described in this paper solves \textsc{The Planar Minimal Branch Decomposition Problem}, which can be computed in polynomial time.

@@

\section{The algorithm}
	This section describes the algorithm given by Seymour and Thomas\cite{ST93} by identifying a set of practical problems and subproblems and how they relate.

@@

@@

	Doing a series of edge contractions (contraction of all edges incident to a pair of vertices) on a graph $M$, where the carving width does not increase until 3 vertices remain, then the series of contracted edges along with the three vertices can be assembled into a minimal carving decomposition of $M$.

@@

	\begin{problem}\label{problem:contraction}
		Given a graph $M$ and a pair of vertices $\{u, v\}$, output a graph resulting from a contraction of all edges incident to $u$ and $v$ if $u$ and $v$ are neighbors.
	\end{problem}

@@

@@

	$E(Q(M, k, e)) = \{e_1 \st $ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

@@

@@

	Let $S$ be every possible state when the rat-catcher is in a face some of which might be losing states. $S = \{ (f, v) \st v \in V(M) \land f$ is a face of $ M \}$.

	Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \st e \in E(M) \land C$ is a component of $Q(M, k, e) \}$.

@@

		As this contraction is a contraction of ALL edges connecting a pair of vertices, the resulting graph will not exhibit any self-loops. I suspect reconciling this and the ordering invariant could be difficult, but luckily in this context, it is irrelevant.

@@

		First, update any edges incident to either $a$ or $b$. Then creating the neighborhood of the new vertex $c$ from the contraction of vertices $a$ and $b$, is done by firstly finding any shared edge $e$. In this implementation the first shared edge $e$ in the neighborhood of $a$. This edge has some ID $e$ and the other half-edge with ID $-e$ will therefore be in the neighborhood of $b$. Now  "rotating" the neighborhoods of $a$ and $b$ such edge $e$ and $-e$ is at index 0 in both lists means that a concatenation of the lists will preserve the ordering around the new vertex $c$. And finally, remove any edges connecting $a$ and $b$.

@@

		Considering a tuple $(f, v)$, if there exists a tuple $(e, C)$ that is losing where $e \in f$ and $v \in V(C)$ then $(f, v)$ is losing.

@@

		\begin{claim}
			The shortest closed walk that includes both $e_1^*$ and $e_2^*$ has the same length as either $$d(s_1, s_2) + d(t_1, t_2) + 2$$ or $$d(s_1, t_1) + d(s_2, t_2) + 2$$. Where $d(u, v)$ is the length of the shortest $u,v$-path.
		\end{claim}

@@

		Using the mapping from links to edges, and the fact that an edge $e$ is called quiet iff. $e$ is not noisy, the quiet edges can be obtained in the natural way.

		Computing the quiet subgraph and the components thereof is done with a depth-first search approach.

		The edges of the components are irrelevant for the rest of the algorithm, so only a list of vertices is returned for each component.

@@

		The dual has a vertex for each face of the input graph. The faces are found by selecting an unmarked half-edge, and then marking all the edges of the face it belongs to, by following the edges that are just next to each other in the ordered neighborhoods.

@@

\newpage
\section{References}
	\printbibliography

\newpage
\section{Appendix}
	\pyfile{branch_decomposition.py}{0}{1000}
	\pyfile{branch_width_brute_force.py}{0}{1000}	
	\pyfile{branch_width.py}{0}{1000}
	\pyfile{carving_decomposition.py}{0}{1000}
	\pyfile{carving_width_brute_force.py}{0}{1000}
	\pyfile{carving_width.py}{0}{1000}
	\pyfile{contraction.py}{0}{1000}
	\pyfile{dual_graph.py}{0}{1000}
	\pyfile{Graph.py}{0}{1000}
	\pyfile{medial_graph.py}{0}{1000}
	\pyfile{parse_graph.py}{0}{1000}
	\pyfile{parse_newick.py}{0}{1000}
